---
title: "HW 2: Energy Demand and CO2 Emissions"
author: "Hayden Atchley"
date: "`r Sys.Date()`"
output:
  pdf_document:
    latex_engine: xelatex
mainfont: Gentium Book Basic
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
	echo = FALSE,
	message = FALSE,
	warning = FALSE
)

library(tidyverse)
library(gridExtra)
library(kableExtra)

options(scipen = 999)
```

# 1

```{r}
vm1 <- read_csv("HW2/data/vm1.csv", n_max = 28) %>% 
	mutate(ITEM = ifelse(is.na(ITEM), lag(ITEM), ITEM))


co2gal <- c(gasoline = 8887, diesel = 10180)
```

Creating a new category combining all buses, single-unit trucks, and combination trucks gives the following:

```{r}
hdv1 <- vm1 %>% 
	mutate("ALL HDV" = BUSES + `SINGLE-UNIT TRUCKS/3` + `COMBINATION TRUCKS`)

vmlong <- hdv1 %>% 
	pivot_longer(c(-YEAR, -ITEM), names_to = "category") %>% 
	arrange(category, ITEM)
```

## 1.1
The average fuel economy is

```{r}
hdvfe <- hdv1 %>% 
	filter(ITEM == "Average miles traveled per  gallon of fuel consumed") %>% 
	select(YEAR, `ALL HDV`) %>% 
	transmute(Year = YEAR, "Fuel economy (mpg)" = `ALL HDV`)

hdvfe %>% 
	kbl(booktabs = TRUE, caption = "Average Fuel Economy of HDVs") %>% 
	kable_styling(latex_options = "HOLD_position")
```

## 1.2
The CO~2~ emissions per mile is given by $\frac{1}{miles/gal} \times \frac{CO_2}{gal} = \frac{CO_2}{mile}$. This gives

```{r}
hdvem <- hdvfe %>% 
	mutate("Fuel consumption (gpm)" = 1/`Fuel economy (mpg)`,
				 "Emissions (g CO$_2$/gal)" = co2gal["diesel"],
				 "Emissions (g CO$_2$/mile)" =
				 	`Fuel consumption (gpm)` * `Emissions (g CO$_2$/gal)`)

hdvem %>% 
	kbl(booktabs = TRUE, escape = FALSE, caption = "HDV Fuel Consumption") %>% 
	kable_styling(latex_options = "HOLD_position")
```

## 1.3
The total CO~2~ emissions for all LDVs and HDVs is

```{r}
allem <- vmlong %>% 
	filter(category %in% c("ALL LIGHT DUTY VEHICLES/2", "ALL HDV"),
				 ITEM %in% c("Fuel consumed   (thousand gallons)")) %>% 
	pivot_wider(names_from = ITEM) %>% 
	mutate("Emissions (g CO$_2$/gal)" = c(rep(co2gal[2], 2),
																				rep(co2gal[1], 2)),
				 category = ifelse(str_detect(category, "LIGHT"),
				 									"ALL LDV", category)) %>% 
	arrange(desc(category), desc(YEAR)) %>% 
	mutate("Emissions (g CO$_2$)" = 
				 	`Fuel consumed   (thousand gallons)` *
				 	`Emissions (g CO$_2$/gal)` * 1000,
				 "Emissions (M tons CO$_2$)" = `Emissions (g CO$_2$)` / 1000000) %>% 
	relocate(category) %>% 
	rename(Year = YEAR, "Vehicle category" = category) 

allem %>% 
	kbl(booktabs = TRUE, escape = FALSE,
			caption = "Emissions of LDVs and HDVs") %>% 
	collapse_rows(columns = 1, valign = "middle") %>% 
	kable_styling(latex_options = c("HOLD_position", "scale_down"))
```

Calculating the percentage of emissions from HDVs per year:

```{r}
pctem <- allem %>% 
	select(Year, `Vehicle category`, `Emissions (M tons CO$_2$)`) %>% 
	pivot_wider(names_from = `Vehicle category`, values_from = `Emissions (M tons CO$_2$)`) %>% 
	transmute(Year,
						"LDV Emissions" = `ALL LDV`,
						"HDV Emissions" = `ALL HDV`,
						"HDV Proportion" = `HDV Emissions` /
							(`HDV Emissions` + `LDV Emissions`))

pctem %>% 
	kbl(booktabs = TRUE, escape = FALSE,
			caption = "HDV Emissions Proportion",
			digits = 3) %>% 
	kable_styling(latex_options = c("HOLD_position"))
```

From this we can see that HDVs made up a greater percentage of CO~2~ emissions in 2020 than in 2019.

## 1.4

Comparing the values I got to the values from the EPS's Greenhouse Gas Emissions Inventory:

```{r}
epaem <- read_csv("HW2/data/epa.csv") %>% 
	filter(`U.S. Emissions, Transportation Sector, MMT CO2 eq.` == "Total") %>% 
	select(`2019`:`2020`)

epaemlist <- epaem %>% as.numeric()
names(epaemlist) <- epaem %>% colnames()

compareem <- pctem %>% 
	select(-`HDV Proportion`) %>%
	mutate("Total Emissions (MMT CO$_2$)" = 
				 	(`LDV Emissions` + `HDV Emissions`) / 1000000) %>% 
	tibble("EPA Values (MMT CO$_2$)" = rev(epaemlist))

compareem %>% 
	kbl(booktabs = TRUE, escape = FALSE,
			caption = "Comparison to EPA Reported Emissions Values") %>% 
	kable_styling(latex_options = c("HOLD_position"))
```

Though the values from my calculations differ from the EPA reported values, they are not wildly different from each other. They are close enough in my judgement that the difference could largely be explained by things like different measuring procedures.



# 2

The drive cycle I'm analyzing is the "Fleet DNA drayage Representative" cycle from NREL. The cycle is shown in Figure \ref{dc}.

```{r, fig.height=4, fig.cap="Fleet DNA drayage Representative drive cycle.\\label{dc}", fig.pos="H"}

	
dc <- read_csv("HW2/data/Fleet DNA Drayage Representative .csv")

dc %>% 
	ggplot(aes(x = `Time (seconds)`, y = `Speed (mph)`)) +
	geom_line(color = "#004466") +
	theme_minimal()
```

I am also using the assumed values in Table \ref{val} for my calculations.

```{r}
CD <- 0.6
AV <- 8.5
mE <- 11000
mC <- 9000
CR <- 0.006
rhoa <- 1.17
ag <- 9.81

mv <- mE + mC

mina <- -3.5
maxa <- 2.5

tribble(
	~Coefficient, ~Symbol, ~Value,
	"Drag coefficient", "$C_D$", "0.6",
	"Frontal area", "$A_v$", "8.5 m$^2$",
	"Empty vehicle weight", "$m_E$", "11,000 kg",
	"Cargo weight", "$m_C$", "9,000 kg",
	"Rolling resistance coefficient", "$C_R$", "0.006",
	"Density of air", "$\\rho_a$", "1.17 kg/m$^3$",
	"Acceleration of gravity", "$g$", "9.81 m/s$^2$"
) %>% 
	kbl(booktabs = TRUE, escape = FALSE,
			caption = "Assumed Values for Calculations\\label{val}",
			linesep = "") %>% 
	kable_styling(latex_options = c("HOLD_position"))
```

## 2.1
The road-load equation for power is given by
\[
P_V = \frac{1}{2}\rho_aC_DA_vv^3 +
v\left(C_Rm_vg + m_vg\sin\alpha + m_va_v\right),
\]
where $m_v = m_E + m_C$, and $\sin\alpha \approx \text{grade (rise/run)}$.

Using this equation, we can calculate the tractive power for each second in the drive cycle. Note that since this is not a hybrid electric vehicle, there is no regenerative braking, and so all negative values of power are set to 0. We also set a limit on acceleration due to speed gaps causing unrealistic values. The limits are $`r mina` < a < `r maxa`$.

```{r}
tp <- dc %>% 
	mutate(
		time = `Time (seconds)`,
		t = time - lag(time),
		v = `Speed (mph)` * 0.44704,
		araw = (v - lag(v)) / t,
		a = case_when(araw < mina ~ mina, araw > maxa ~ maxa, TRUE ~ araw),
		grade = `Grade (rise/run)`,
		
		dpowerraw = (1/2 * rhoa*CD*AV*v^3) / 1000,
		dpower = ifelse(dpowerraw < 0, 0, dpowerraw),
		rpowerraw = v*CR*mv*ag / 1000,
		rpower = ifelse(rpowerraw < 0, 0, rpowerraw),
		apowerraw = v*mv*a / 1000,
		apower = ifelse(apowerraw < 0, 0, apowerraw),
		gpowerraw = v*mv*ag*grade / 1000,
		gpower = ifelse(gpowerraw < 0, 0, gpowerraw),
		
		powerraw = dpowerraw + apowerraw + rpowerraw + gpowerraw,
		power = ifelse(powerraw < 0, 0, powerraw),
		
		energy = power*t/3600, #kWh
		
		dist = v*t
		) %>% 
	select(-c(`Time (seconds)`, `Speed (mph)`, `Grade (rise/run)`)) %>% 
	relocate(time, v)
```

The plot of the tractive power is given in Figure \ref{tp}.

```{r, fig.cap="Plot of tractive power for drayage drive cycle.\\label{tp}", fig.height=4}
tp %>% 
	ggplot(aes(x = time, y = power)) +
	geom_line(color = "#660044") +
	theme_minimal() +
	labs(x = "Time (seconds)",
			 y = "Tractive Power (kW)")
```

## 2.2
```{r}
timemin <- 13900
timemax <- 14000
```

A graph of engine power along with tractive power is given in Figure \ref{powercomp}. I am using only from `r timemin` to `r timemax` seconds for clarity of the graph.

```{r, fig.height=4, fig.cap="Comparison of engine and tractive power.\\label{powercomp}"}
powercomp <- tp %>% 
	filter(time <= timemax, time >= timemin) %>% 
	transmute(time, power, engine = `Engine Power (kW)`,
						diff = power - engine) %>% 
	pivot_longer(-time) %>% 
	mutate(name = as.factor(name))

powercomp$name <- powercomp$name %>%  
	fct_relevel(c("power", "engine", "diff"))

powercomp %>% 
	ggplot(aes(x = time, y = value, color = name)) +
	geom_line() +
	labs(x = "Time (seconds)",
			 y = "Power (kW)") +
	scale_color_discrete(name = "", labels = c(
		diff = "Tractive minus Engine", engine = "Engine Power", power = "Tractive Power")) +
	theme_minimal() + 
  theme(legend.position = 'bottom')
```

The data generally seem to make sense, as the engine power is often more than the tractive power, which is expected due to losses. Though I am not sure why tractive power sometimes exceeds engine power; it appears to have something to do with the spikes in tractive power, which might be due to some time gaps in the data, or possibly a small temporal offset between the engine and tractive power.

## 2.3
The total energy required to move the vehicle for this drive cycle is `r sum(tp$energy, na.rm=TRUE) %>% round()` kWh.

## 2.4
```{r}
drag <- sum(tp$dpower*tp$t/3600, na.rm = TRUE)
roll <- sum(tp$rpower*tp$t/3600, na.rm = TRUE)
accel <- sum(tp$apower*tp$t/3600, na.rm = TRUE)
grav <- sum(tp$gpower*tp$t/3600, na.rm = TRUE)

total <- sum(drag, roll, accel, grav)


tibble(
	" " = c("Energy (kWh)", "%"),
	Drag = c(drag, drag/total * 100),
	Rolling = c(roll, roll/total * 100),
	Acceleration = c(accel, accel/total * 100),
	Gravity = c(grav, grav/total * 100),
	Total = c(total, total/total * 100)
) %>% 
	kbl(booktabs = TRUE, digits = 1, align = 'r',
			caption = "Components of Total Tractive Power\\label{comp}") %>% 
	kable_styling(latex_options = c("HOLD_position"))
```

## 2.5
```{r}
regen <- -sum(tp$powerraw[tp$powerraw < 0] * 
								tp$t[tp$powerraw < 0]/3600, na.rm = TRUE)

oldpower <- sum(tp$energy, na.rm = TRUE)
newpower <- oldpower - (regen/2)
```

Assuming the "negative" tractive power would be used for regenerative braking, an additional `r (regen) %>% round()` kWh would be available. If 50% of that power was recaptured, then the additional tractive power needed would be reduced to `r newpower %>% round()` kWh, a reduction of `r ((oldpower - newpower)/oldpower * 100) %>% round(2)`%.

## 2.6
```{r}
enginemove <- tp %>% 
	mutate(newengine = `Engine Power (kW)` - 7,
				 newengine = ifelse(newengine < 0, 0, newengine))

moveenergy <- sum(enginemove$newengine*enginemove$t/3600, na.rm = TRUE)
```

Assuming an auxiliary power load of 7 kW, the engine power used to move the vehicle is $P_{Engine} - 7$ kW, but not less than 0 kW. Finding the total energy expended by the engine attempting to move the vehicle gives `r moveenergy %>% round()` kWh. The total energy actually required to move the vehicle is `r (oldpower) %>% round()` kWh, so the efficiency of the drivetrain is `r (oldpower/moveenergy * 100) %>% round(1)`%.

## 2.7
Low-sulfur diesel has an energy content of 128,488 Btu/gal according to the AFDC. This is equivalent to `r (energy <- 37.656)` kWh/gal. Assuming an engine efficiency of `r (eff <- 0.3)`, the total fuel needed for this drive cycle is $F = \frac{`r (oldpower) %>% round()`[\text{kWh}]/`r eff`}{`r energy %>% round(1)`[\text{kWh/gal}]} = `r (fuel1 <- oldpower/energy/eff) %>% round()`\ \text{gal}$. The total distance in this drive cycle is `r (totdist <- sum(tp$dist, na.rm = TRUE)) %>% round()` meters or `r (miles <- (totdist * .0006213712)) %>% round(1)` miles, giving an overall fuel economy of `r (miles/fuel1) %>% round(2)` mpg, and a fuel consumption of `r (fuel1/miles*1000) %>% round()` gallons per 1000 miles.

## 2.8
The energy required in the regenerative-braking scenario (2.5) is `r newpower %>% round()` kWh. The fuel needed is now $F = \frac{`r (newpower) %>% round()`[\text{kWh}]/`r eff`}{`r energy %>% round(1)`[\text{kWh/gal}]} = `r (fuel2 <- newpower/energy/eff) %>% round()`\ \text{gal}$, with a fuel economy and consumption of `r (miles/fuel2) %>% round(2)` mpg and `r (fuel2/miles*1000) %>% round()` gallons per 1000 miles. If this drive cycle (`r (miles <- (totdist * .0006213712)) %>% round(1)` miles) is representative of a typical day, then at 5 days per week and 52 weeks per year, the fuel saved is $5 \text{[days/wk]} \times 52 \text{[wk/yr]} \times (`r fuel1 %>% round()` - `r fuel2 %>% round()`) \text{[gal]} = `r (fuelsave <- 5*52*(fuel1-fuel2)) %>% round` \text{[gal/yr]}$ for each truck. Given a current (26 Sept. 2022) diesel price of \$`r (price <- 4.889)`/gal, this amounts to a savings of \$`r (price*fuelsave) %>% round()` per year per truck.



# 3
From Annual Energy Outlook 2022 Table 40, we get that the 2021 fuel economies of certain vehicle technologies are as follows:

```{r}
gasmpg <- 42.94
hybmpg <- 62.51
evmpg <- 101.9

kwhtogal <- 0.03

coalem <- 767/757763 * 1000
ngem <- 576/1402438 * 1000
```

- ICE vehicles: `r gasmpg` mpg
- Hybrid vehicles: `r hybmpg` mpge
- 200 mile range EVs: `r evmpg` mpge

From the EIA we obtain the emission rate of coal and natural gas energy production:

```{r}
emissions <- tibble(
	"Energy Source" = c("Coal", "Natural Gas", "Petroleum"),
	electricity = c(757763, 1402438, 13665),
	co2 = c(767, 576, 13),
	emrate = co2 / electricity * 1000
	)

emissions %>% 
	rename(
		"Electricity (M kWh)" = electricity,
		"CO$_2$ Emissions (M MT)" = co2,
		"CO$_2$ per kWh (kg)" = emrate
	) %>% 
	kbl(digits = 3,
			caption = "CO$_2$ Emissions per Kilowatt-hour Produced",
			escape = FALSE,
			booktabs = TRUE) %>% 
	kable_styling(latex_options = c("HOLD_position"))

emc <- emissions$co2[1]
emng <- emissions$co2[2]

elecc <- emissions$electricity[1]
elecng <- emissions$electricity[2]

emratep <- emissions$emrate[3]

weightem <- ((emc + emng) / (elecc + elecng)) * 1000
```

Assuming the electricity used in the EVs is produced from coal and natural gas, we can calculate an average of the two emission rates. By summing the total emissions and total energy generated, we find that the overall emission rate is $\frac{`r emc` + `r emng`}{`r elecc` + `r elecng`} \times 1000 = `r weightem %>% round(2)`$ kg CO~2~ per kWh.

## 3.1
The following table shows information on CO~2~ emissions by vehicle type. I am assuming that the emission rate from petroleum electricity generation is similar enough to the rate when using an ICE vehicle to use in these calculations. I am also using a conversion factor of 1 kWh to `r kwhtogal` gallons of gasoline.

```{r}
emtable <- tibble(
	"Vehicle Type" = c("ICE", "Hybrid", "EV"),
	mpg = c(gasmpg, hybmpg, evmpg),
	mpkwh = mpg*kwhtogal,
	co2pkwh = c(emratep, emratep, weightem),
	co2pmi = co2pkwh / mpkwh * 1000
)

emtable %>% 
	rename(
		"MPG (equiv)" = mpg,
		"Miles per kWh" = mpkwh,
		"CO$_2$ per kWh (kg)" = co2pkwh,
		"CO$_2$ per Mile (g)" = co2pmi
	) %>% 
	kbl(digits = c(0,1,2,3,0),
			caption = "CO$_2$ Emissions per Mile by Vehicle Type",
			escape = FALSE,
			booktabs = TRUE) %>% 
	kable_styling(latex_options = c("HOLD_position"))
```

The EV has the lowest emission rate per mile, while the ICE vehicle has the highest. This follows my intuition, as even with the inefficiencies in electricity production EVs are much more efficient per kWh than gas-powered vehicles. The emission rate of the EVs could theoretically fall even further if cleaner electricity sources are used.

## 3.2
Using the EPA 2020 average of 25.5 mpg and 350 g CO~2~/mile for the ICE vehicle, the hybrid vehicle now has the worst emission rate. Though in reality this probably would not be the case; a lower emission rate per gallon of gasoline used would benefit both ICE and hybrid vehicles. It's possible my previous assumption that emissions would be similar between an ICE and petroleum electricity generation is misguided, in which case the numbers I calculated may change significantly.
